---
title: "W271 HW6"
author: "Subhashini R., Lei Yang, Ron Cordell"
date: "March 16, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1:

**a. Discuss the mean and variance functions and how the similarities and differences from those we studied in classical linear model**

The Mean Function for a time series is given by:

$$u_{x}(t) = E(x_{t}) = \int_{-\infty}^{\infty}x_{t}f_{t}(x_{t})dx_{t}$$

The mean function is dependent on time $t$ and the expectation is over the ensemble of time series $x_{t}$ derived from the underlying process.

A process is stationary in the mean if the mean function is a constant in time.

The Variance Function for a time series is given by:

$$\sigma_{x}^{2}(t) = E(x_{t}-\mu_{x}(t))^{2} = \int_{-\infty}^{\infty}(x_{t}-\mu_{x}(t))^{2}f_{t}(x_{t})dx_{t}$$

Once again the expectation is for the ensemble of all time series.

A series is stationary in variance if ther variance function is constant in time.

**b. Define strict and weak stationarity**

A time series is strictly stationary if the joint distributions $F(x_{t_{1}},...,x_{t_{n}})$ and $F(x_{t_{1}+m},...,x_{t_{n}+m})$ are the same $\forall t_{1},...,t_{n}$ and $m$. In other words, the distribution is the same regardless of the time period.

A time series is weakly stationary if it is mean and variance stationary and the autocovariance $Cov(x_{t},x_{t+k})$ depends on the time placement $k$.

# Exercise 2:

**a. Generate a zero-drift random walk model using 500 simulation**  
**b. Provide the descriptive statistics of the simulated realizations. The descriptive statistics should include the mean, standard deviation, 25th, 50th, and 75th quantiles, minimum, and maximum**  
**c. Plot the time-series plot of the simulated realizations**  
**d. Plot the autocorrelation graph**  
**e. Plot the partial autocorrelation graph** 

```{r}
# set seed for repeatability
set.seed(1)

# 500 random normal "draws"
x1 <- w <- rnorm(500)

# create the lagged variable
for (i in 2:500) x1[i] <- x1[i-1] + w[i]

# let's see what we've got
str(x1)
str(w)

# summary stats
summary(x1)
sd(x1)

# put our graphs in the same page/display canvas
plot.ts(x1, col='blue', 
        xlab='Simulated Time Period',
        ylab='Simulated Variable Units')
title('(2c) Fig 1: Random Walk Zero Drift Simulated Series')
```
```{r}
acf(x1, main="", 
    xlab='Lag')
title('(2d) Fig 2: ACF of the Random Walk Zero Drift Simulated Series')
```
```{r}
pacf(x1, main="", 
     xlab='Lag')
title('(2e) Fig 3: PACF of the Random Walk Zero Drift Simulated Series')

```

# Exercise 3:

**a. Generate a random walk with drift model using 500 simulation, with the drift = 0.5**  
**b. Provide the descriptive statistics of the simulated realizations. The descriptive statistics should include the mean, standard deviation, 25th, 50th, and 75th quantiles, minimum, and maximum**  
**c. Plot the time-series plot of the simulated realizations**  
**d. Plot the autocorrelation graph**  
**e. Plot the partial autocorrelation graph**  

```{r}
# we'll use our same series w
x2 <- w

# set the drift to 0.5
drift <- 0.5

# calculated the lagged variable with the drift
for (i in 2:500) x2[i] <- drift + x2[i-1] + w[i]

# summary stats
summary(x2)
sd(x2)

# put our graphs in the same page/display canvas
plot.ts(x2, col='blue', 
        xlab='Simulated Time Period',
        ylab='Simulated Variable Units')
title('(3c) Fig 1: Random Walk 0.5 Drift Simulated Series')
```
```{r}
acf(x1, main="", 
    xlab='Lag')
title('(3d) Fig 2: ACF of the Random Walk 0.5 Drift Simulated Series')
```
```{r}
pacf(x1, main="", 
     xlab='Lag')
title('(3e) Fig 3: PACF of the Random Walk 0.5 Drift Simulated Series')
```

# Exercise 4:

Use the series from INJCJC.csv

**a. Load the data and examine the basic structure of the data using str(), dim(), head(), and tail() functions**  

```{r}
# load the INJCJC.csv data into a dataframe
df <- read.csv('INJCJC.csv')

# examine the structure
str(df)

# examine the dimensions of the data set
dim(df)

# examine the first 20 rows
head(df,20)

#examine the last 20 rows
tail(df,20)
```

**b. Convert the variables INJCJC into a time series object frequency=52, start=c(1990,1,1), end=c(2014,11,28). Examine the converted data series**  

```{r}
INJCJC <- ts(df$INJCJC, frequency=52, start=c(1990,1,1), end=c(2014,11,28))
summary(INJCJC)
str(INJCJC)
```

**c. Define a variable using the command INJCJC.time<-time(INJCJC)**  

```{r}
INJCJC.time <- time(INJCJC)
str(INJCJC)
```

**d. Using the following command to examine the first 10 rows of the data. Change the parameter to examine different number of rows of data**

    head(cbind(INJCJC.time, INJCJC),10)
    
```{r}
head(cbind(INJCJC.time, INJCJC), 20)
tail(cbind(INJCJC.time, INJCJC), 20)
```

**e1. Plot the time series plot of INJCJC. Remember that the graph must be well labelled.**  

```{r}
plot.ts(INJCJC, col='blue',
        xlab='Year (Weekly Interval)',
        ylab='Price',
        main='Plot of Time Series INJCJC')
```

**e2. Plot the histogram of INJCJC. What is shown and not shown in a histogram? How do you decide the number of bins used?**

```{r}
hist(INJCJC, breaks=80, main='Histogram of Time Series INJCJC',
     xlab='INJCJC Price')
```

The histogram shows the relative frequency of values in the series, but it does not show the time dependency of those values, and therefore their relationship to each other as a function of time. I decide the number of bins to use to show a reasonable amount of detail in the variation of the histogram while giving a good indication of the distribution of values.

**e3. Plot the autocorrelation graph of INJCJC series**  

```{r}
acf(INJCJC, main='ACF of the INJCJC Time Series')

```

**e4. Plot the partial autocorrelation graph of INJCJC series** 

```{r}
pacf(INJCJC, main='ACF of the INJCJC Time Series')
```
\newpage

**e5. Plot a 3x3 Scatterplot Matrix of correlation against lag values**  

```{r fig.width=6.5, fig.height=8}
lag.plot(INJCJC, 9, main='Correlation vs. Lag Plot Matrix of INJCJC')
```
\newpage

**f1. Generate two symmetric Moving Average Smoothers. Choose the number of moving average terms such that one of the smoothers is very smooth and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.**

```{r fig.width=6.5, fig.height=8}
# The time series is weekly, we we can try a moving window 
# average of 5 that would smooth across months
INJCJC.ma5 <- filter(INJCJC, sides=2, rep(1,5)/5)

# This symmetric MA filter would smooth across 2 years - 
# the 1 year MA smoother is much less smooth
INJCJC.ma105 <- filter(INJCJC, sides=2, rep(1,105)/105)
plot(INJCJC, type='p', pch=20, col='black', xlab='Year',
     main='INJCJC Time Series Symmetric Moving Average', ylab='Price')
lines(INJCJC.ma5, col='cyan', lty=1, lwd=1.5)
lines(INJCJC.ma105, col='red', lty=1, lwd=2.0)
legend('topleft',legend=c('Time Series',
                          'Moving Average (5) Monthly Smoothed',
                          'Moving Average (105) 2-Year Smoothed'),
       lty=c(1,1,1), col=c('black','cyan','red'), bty='n', cex=1,
       merge=TRUE, bg=336)
```
\newpage

**f2. Generate two regression smoothers, one being a cubic trend regression and the other being a periodic regression. Plot the smoothers and the original series in one graph.**

```{r fig.width=6.5, fig.height=8}
wk <- INJCJC.time - mean(INJCJC.time)
wk2 <- wk^2
wk3 <- wk^3
cs <- cos(.29*pi*wk)
sn <- sin(.29*pi*wk)
INJCJC.cubic <- lm(INJCJC ~ wk + wk2 + wk3, na.action=NULL)
INJCJC.periodic <- lm(INJCJC ~ wk + wk2 + wk3 + cs + sn, na.action=NULL)
plot(INJCJC, type='p', pch=20, col='black', xlab='Year',
     main='INJCJC Time Series Cubic Polynomial and Periodic Smoothing', ylab='Price')
lines(fitted(INJCJC.cubic), lty=1, lwd=1.5, col='cyan')
lines(fitted(INJCJC.periodic), lty=1, lwd=2.0, col='red')
legend('topleft',legend=c('Time Series',
                          'Cubic Regression Smoothed',
                          'Periodic Smoothed'),
       lty=c(1,1,1), col=c('black','cyan','red'), bty='n', cex=1,
       merge=TRUE, bg=336)
```
\newpage

**f3. Generate kernel smoothers. Choose the smoothing parametrs such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.**

```{r fig.width=6.5, fig.height=8}
plot(INJCJC, type='p', pch=20, col='black', xlab='Year',
     main='INJCJC Time Series Kernel Smoothing', ylab='Price')
lines(ksmooth(time(INJCJC), INJCJC, "normal", bandwidth=5/52), lty=1, lwd=1.5, col='cyan')
lines(ksmooth(time(INJCJC), INJCJC, "normal", bandwidth=2), lty=1, lwd=2.0, col='red')
legend('topleft',legend=c('Time Series',
                          'Bandwidth 5/52',
                          'Bandwidth 2'),
       lty=c(1,1,1), col=c('black','cyan','red'), bty='n', cex=1,
       merge=TRUE, bg=336)
```
\newpage

**f4. Generate two nearest neighborhood smoothers. Choose the smoothing parametrs such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.**


```{r fig.width=6.5, fig.height=8}
plot(INJCJC, type='p', pch=20, col='black', xlab='Year',
     main='INJCJC Time Series Nearest Neighbor Smoothing', ylab='Price')
lines(supsmu(time(INJCJC), INJCJC, span=0.01), lty=1, lwd=1.5, col='cyan')
lines(supsmu(time(INJCJC), INJCJC, span=0.2), lty=1, lwd=2.0, col='red')
legend('topleft',legend=c('Time Series',
                          'Span 0.01',
                          'Span 0.2'),
       lty=c(1,1,1), col=c('black','cyan','red'), bty='n', cex=1,
       merge=TRUE, bg=336)
```
\newpage

**f5. Generate two LOWESS smoothers. Choose the smoothing parametrs such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.**



```{r fig.width=6.5, fig.height=8}
plot(INJCJC, type='p', pch=20, col='black', xlab='Year',
     main='INJCJC Time Series Lowess Smoothing', ylab='Price')
lines(lowess(time(INJCJC), INJCJC, f=0.02), lty=1, lwd=1.5, col='cyan')
lines(lowess(time(INJCJC), INJCJC, f=0.25), lty=1, lwd=2.0, col='red')
legend('topleft',legend=c('Time Series',
                          'Span 0.01',
                          'Span 0.2'),
       lty=c(1,1,1), col=c('black','cyan','red'), bty='n', cex=1,
       merge=TRUE, bg=336)
```
\newpage

**f6. Generate two spline smoothers. Choose the smoothing parametrs such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.**

```{r fig.width=6.5, fig.height=8}
plot(INJCJC, type='p', pch=20, col='black', xlab='Year',
     main='INJCJC Time Series Spline Smoothing', ylab='Price')
lines(smooth.spline(time(INJCJC), INJCJC, spar=0.05), lty=1, lwd=1.5, col='cyan')
lines(smooth.spline(time(INJCJC), INJCJC, spar=0.75), lty=1, lwd=2.0, col='red')
legend('topleft',legend=c('Time Series',
                          'Spar 0.01',
                          'Spar 0.2'),
       lty=c(1,1,1), col=c('black','cyan','red'), bty='n', cex=1,
       merge=TRUE, bg=336)
```