---
title: "W271 - Applied Regression and Time Series Analysis - HW6"
author: "Subhashini R., Lei Yang, Ron Cordell"
date: "March 17, 2016"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Exercise 1:
a). Discuss the mean and variance functions and how the similarities and differences from those we studied in classical linear model

b). Define strict and weak statonarity

# Exercise 2:
a). Generate a zero-drift random walk model using 500 simulation

```{r}
# white noise with sample size 500
w <- rnorm(500,0,1)
# random walk
rw <- cumsum(w)
```

b). Provide the descriptive statistics of the simulated realizations. The descriptive statistics should include the mean, standard deviation, 25th, 50th, and 75th quantiles, minimum, and maximum

```{r echo=FALSE}
Q <- quantile(rw, probs=c(0.25, 0.5, 0.75))
```

| Statistic        | Value      |
| ------------- |-------------:|
| mean    | **``r sprintf("%.4f", mean(rw))``** | 
| standard deviation      | **``r sprintf("%.4f", sd(rw))``**      |
| 25% Q | **``r sprintf("%.4f", Q[1])``**  |
| 50% Q | **``r sprintf("%.4f", Q[2])``**  |
| 75% Q | **``r sprintf("%.4f", Q[3])``**  |
| minimum | **``r sprintf("%.4f", min(rw))``**  |
| maximum | **``r sprintf("%.4f", max(rw))``**  |

c). Plot the time-series plot of the simulated realizations
```{r}
plot.ts(rw, ylab='Simulated Value', main='Zero-drift Random Walk')
```

d). Plot the autocorrelation graph
```{r}
acf(rw, main="Correlogram of Zero-drift Random Walk")

```

e). Plot the partial autocorrelation graph
```{r}
pacf(rw, main="Partial Autocorrelation of Zero-drift Random Walk")

```

# Exercise 3:
a). Generate arandom walk with drift model using 500 simulation, with the drift = 0.5
```{r}
w <- rnorm(500,0,1)+0.5
rw <- cumsum(w)
```

b). Provide the descriptive statistics of the simulated realizations. The descriptive statistics should include the mean, standard deviation, 25th, 50th, and 75th quantiles, minimum, and maximum
```{r echo=FALSE}
# descriptive stats

Q <- quantile(rw, probs=c(0.25, 0.5, 0.75))

```

| Statistic        | Value      |
| ------------- |-------------:|
| mean    | **``r sprintf("%.4f", mean(rw))``** | 
| standard deviation      | **``r sprintf("%.4f", sd(rw))``**      |
| 25% Q | **``r sprintf("%.4f", Q[1])``**  |
| 50% Q | **``r sprintf("%.4f", Q[2])``**  |
| 75% Q | **``r sprintf("%.4f", Q[3])``**  |
| minimum | **``r sprintf("%.4f", min(rw))``**  |
| maximum | **``r sprintf("%.4f", max(rw))``**  |

c). Plot the time-series plot of the simulated realizations
```{r}
plot.ts(rw, ylab='Simulated Value', main='0.5-drift Random Walk')
```

d). Plot the autocorrelation graph
```{r}
acf(rw, main="Correlogram of 0.5-drift Random Walk")
```

e). Plot the partial autocorrelation graph
```{r}
pacf(rw, main="Partial Autocorrelation of 0.5-drift Random Walk")
```

# Exercise 4:
Use the series from INJCJC.csv

a). Load the data and examine the basic structure of the data using str(), dim(), head(), and tail() functions
```{r}
setwd("~/Desktop/W271Data")
INJCJC <- read.csv('INJCJC.csv')
str(INJCJC)
dim(INJCJC)
head(INJCJC)
tail(INJCJC)
```

b). Convert the variables INJCJC into a time series object frequency=52, start=c(1990,1,1), end=c(2014,11,28). Examine the converted data series
```{r}
injcjc <- ts(data=INJCJC$INJCJC, frequency=52, start=c(1990,1,1), end=c(2014,11,28))
injcjc4 <- ts(data=INJCJC$INJCJC4, frequency=52, start=c(1990,1,1), end=c(2014,11,28))
#plot.ts(injcjc4, main='', ylab='INJCJC4')
```

c). Define a variable using the command INJCJC.time<-time(INJCJC)
```{r}
INJCJC.time<-time(injcjc)
```

d). Using the following command to examine the first 10 rows of the data. Change the parameter to examine different number of rows of data
```{r}
head(cbind(INJCJC.time, injcjc),18)
```

e1). Plot the time series plot of INJCJC. Remember that the graph must be well labelled.
```{r}
plot.ts(injcjc, main='', ylab='INJCJC')
```

e2). Plot the histogram of INJCJC. What is shown and not shown in a histogram? How do you decide the number of bins used?
```{r}
hist(injcjc, breaks=20)
```

the histogram contains no temperal information in the series, it is impossible to know the dynamics over time from histogram

e3). Plot the autocorrelation graph of INJCJC series
```{r}
acf(injcjc, main='Autocorrelation graph of INJCJC')
```


e4). Plot the partial autocorrelation graph of INJCJC series
```{r}
pacf(injcjc, main='Partial autocorrelation graph of INJCJC')
```

e5). Plot a 3x3 Scatterplot Matrix of correlation against lag values
```{r warning=FALSE}
lag.plot(injcjc, lags=9, layout=c(3,3), 
         diag=TRUE, disg.col="red",
         main="Autocorrelation between INJCJC and its Own Lags")
```


f1). Generate two symmetric Moving Average Smoothers. Choose the number of moving average terms such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.
```{r}
ma3 = filter(injcjc, sides=2, rep(1,18)/18)
ma18 = filter(injcjc, sides=2, rep(1,118)/118)
plot(injcjc, main="", pch=4, lty=5, lwd=1, xlab="Year")
lines(ma3, lty=1, lwd=1.5, col="green")
lines(ma18, lty=1, lwd=1.5, col="blue")
leg.txt <- c("Original Series", "18-Point Symmetric Moving Average", "118-Point Symmetric Moving Average")
legend("topleft", legend=leg.txt, lty=c(1,1,1), col=c("black","green","blue"),
       bty='n', cex=1, merge = TRUE, bg=336)
```


f2). Generate two regression smoothers, one being a cubic trend regression and the other being a periodic regression. Plot the smoothers and the original series in one graph.
```{r}
wk = time(injcjc) - mean(time(injcjc))  
wk2 = wk^2 
wk3 = wk^3
cs = cos(0.8*pi*wk)  
sn = sin(0.8*pi*wk)
reg1 = lm(injcjc~wk + wk2 + wk3, na.action=NULL)
reg2 = lm(injcjc~wk + wk2 + wk3 + cs + sn, na.action=NULL)
plot(injcjc, main="Regression Smoothing", 
     pch=4, lty=5, lwd=1, xlab="Year")
lines(fitted(reg1), lty=1, lwd=1.5, col="green")
lines(fitted(reg2), lty=1, lwd=1.5, col="blue")
# Add Legend
leg.txt <- c("Original Series", "Cubic Trend Regression Smoothing", "Periodic Regression Smoothing")
legend("topleft", legend=leg.txt, lty=c(1,1,1), col=c("black","green","blue"),
       bty='n', cex=1, merge = TRUE, bg=336)
```


f3). Generate kernel smoothers. Choose the smoothing parametrs such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.
```{r}
plot(injcjc, pch=4, lty=5, lwd=1, xlab="Year")
lines(ksmooth(time(injcjc), injcjc, "normal", bandwidth=5/104),lty=1, lwd=1.5, col="green")
lines(ksmooth(time(injcjc), injcjc, "normal", bandwidth=2),lty=1, lwd=1.5, col="blue")
# Add Legend
leg.txt <- c("Original Series", "Kernel Smoothing: bandwidth=5/104", "Kernel Smoothing: bandwidth=2")
legend("topleft", legend=leg.txt, lty=c(1,1,1), col=c("black","green","blue"),
       bty='n', cex=1, merge = TRUE, bg=336)

```


f4). Generate two nearest neighborhood smoothers. Choose the smoothing parametrs such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.
```{r}
plot(injcjc, main="Nearest Neighborhood Smoothing", 
     pch=4, lty=5, lwd=1, xlab="Year")
lines(supsmu(time(injcjc), injcjc, span=.01),lty=1, lwd=1.5, col="green")
lines(supsmu(time(injcjc), injcjc, span=.2),lty=1, lwd=1.5, col="blue")
# Add Legend
leg.txt <- c("Original Series", "NN Smoothing: bandwidth=.01", "NN Smoothing: bandwidth=.2")
legend("topleft", legend=leg.txt, lty=c(1,1,1), col=c("black","green","blue"),
       bty='n', cex=1, merge = TRUE, bg=336)
```

f5). Generate two LOWESS smoothers. Choose the smoothing parametrs such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.
```{r}
plot(injcjc, main="LOWESS Smoothing", pch=4, lty=5, lwd=1, xlab="Year")
lines(lowess(injcjc, f=.02),lty=1, lwd=1.5, col="green")
lines(lowess(injcjc, f=.23),lty=1, lwd=1.5, col="blue")
# Add Legend
leg.txt <- c("Original Series", "LOWESS Smoothing: bandwidth=.02", "LOWESS Smoothing: bandwidth=.23")
legend("topleft", legend=leg.txt, lty=c(1,1,1), col=c("black","green","blue"),
       bty='n', cex=1, merge = TRUE, bg=336)
```

f6). Generate two spline smoothers. Choose the smoothing parametrs such that one of the smoothers is very smoother and the other one can trace through the dynamics of the series. Plot the smoothers and the original series in one graph.
```{r}
plot(injcjc, main="Smoothing Splines", pch=4, lty=5, lwd=1, xlab="Year")
lines(smooth.spline(time(injcjc), injcjc, spar=0.05),lty=1, lwd=1.5, col="green")          
lines(smooth.spline(time(injcjc), injcjc, spar=1.0),lty=1, lwd=1.5, col="blue")  
# Add Legend
leg.txt <- c("Original Series", "Spline: Smoothing Parameter=.05", "Spline: Smoothing Parameter=1.0")
legend("topleft", legend=leg.txt, lty=c(1,1,1), col=c("black","green","blue"),
       bty='n', cex=1, merge = TRUE, bg=336)
```

