---
title: "W271 HW8"
author: "Subhashini R., Lei Yang, Ron Cordell"
date: "March 29, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Build an univariate linear time series model (i.e AR, MA, and ARMA models) using the series in hw08_series.csv.  

* Use all the techniques that have been taught so far to build the model, including date examination, data visualization, etc.  
* All the steps to support your final model need to be shown clearly.  
* Show that the assumptions underlying the model are valid.  
* Which model seems most reasonable in terms of satisfying the model’s underling assumption?  
* Evaluate the model performance (both in- and out-of-sample)  
* Pick your “best” models and conduct a 12-step ahead forecast. Discuss your results. Discuss the choice
of your metrics to measure “best”.  

```{r}
# Load the libraries and tools
library(astsa)
library(zoo)
library(forecast)
library(stargazer)
```

```{r}
# load the CSV file
df <- read.csv('hw08_series.csv')
str(df)
```

The CSV file for the HW8 time series consists of two variables: an X variable that is the time interval and an x value corresponding to the time period. There is no information about the time interval or units of the values. 

A time series object is created from the dataframe for further analysis.

``` {r}
ts1 <- ts(df$x)
str(ts1)
summary(ts1)
head(ts1)
tail(ts1)
```
\newpage

```{r fig.height=8,fig.width=6}
par(mfrow=c(2,2))
hist(ts1, main='Series Histogram', xlab='x', breaks=20) 
plot.ts(ts1, col='blue',
        xlab='Time Interval',
        ylab='x',
        main='HW8 Time Series')
acf(ts1, main='Autocorrelation')
pacf(ts1, main='Partial Autocorrelation')
```

The time series plot reveals that the HW8 time series is a persistently upward trending series and is not stationary. The autocorrelation shows a very long decay over more than 25 lags while the partial autocorrelation shows statistically significant results at lags 13 and 25.

The ACF plot is indicative of an ARMA underlying model because of the long taper - most likely an ARMA(2,1) model.

```{r}
ts1.fit <- arima(ts1, order=c(2,0,1))
ts1.fit
summary(ts1.fit)
```

The coefficients of the ARMA(2,1) model are all statistically significant.

Checking the residuals...

```{r fig.height=8,fig.width=6}
par(mfrow=c(2,2))
hist(ts1.fit$residuals)
plot.ts(ts1.fit$residuals)
acf(ts1.fit$residuals)
pacf(ts1.fit$residuals)
```

These residuals show a definite trend in that they become more volatile and larger over time. The distribution is slightly skewed. The autocorrelation shows correlations at lags 12 and 24 while the partial autocorrelation shows statistically significant effects at lag 2, 9-14 and 24.

```{r}
plot.ts(ts1)
plot.ts(ts1.fit$series)
plot.ts(ts1.fit$residuals)

```

