---
title: "W271 HW8"
author: "Subhashini R., Lei Yang, Ron Cordell"
date: "March 29, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Build an univariate linear time series model (i.e AR, MA, and ARMA models) using the series in hw08_series.csv.  

* Use all the techniques that have been taught so far to build the model, including date examination, data visualization, etc.  
* All the steps to support your final model need to be shown clearly.  
* Show that the assumptions underlying the model are valid.  
* Which model seems most reasonable in terms of satisfying the model’s underling assumption?  
* Evaluate the model performance (both in- and out-of-sample)  
* Pick your “best” models and conduct a 12-step ahead forecast. Discuss your results. Discuss the choice
of your metrics to measure “best”.  

```{r}
# Load the libraries and tools
library(astsa)
library(zoo)
library(forecast)
library(stargazer)
```

```{r}
# load the CSV file
df <- read.csv('hw08_series.csv')
str(df)
```

The CSV file for the HW8 time series consists of two variables: an X variable that is the time interval and an x value corresponding to the time period. There is no information about the time interval or units of the values. 

A time series object is created from the dataframe for further analysis.

``` {r}
ts1 <- ts(df$x)
str(ts1)
summary(ts1)
head(ts1)
tail(ts1)
```
\newpage

```{r fig.height=8,fig.width=6}
par(mfrow=c(2,2))
hist(ts1, main='Series Histogram', xlab='x', breaks=20) 
plot.ts(ts1, col='blue',
        xlab='Time Interval',
        ylab='x',
        main='HW8 Time Series')
acf(ts1, main='Autocorrelation')
pacf(ts1, main='Partial Autocorrelation')
```

The time series plot reveals that the HW8 time series is a persistently upward trending series and is not stationary. The autocorrelation shows a very long decay over more than 25 lags while the partial autocorrelation shows statistically significant results at lags 13 and 25, indicating a strong seasonal component that happens every 12 periods, in addition to the inter-period seasonality.

The ACF plot is indicative of an ARMA underlying model because of the long taper - most likely an ARMA(2,1) model.

```{r}
ts1.fit <- Arima(ts1, order=c(2,0,1))
ts1.fit
summary(ts1.fit)
```

The coefficients of the ARMA(2,1) model are all statistically significant.

Checking the residuals...

```{r fig.height=8,fig.width=6}
par(mfrow=c(2,2))
hist(ts1.fit$residuals)
plot.ts(ts1.fit$residuals)
acf(ts1.fit$residuals)
pacf(ts1.fit$residuals)
```

These residuals show a definite trend in that they become more volatile and larger over time. The distribution is slightly skewed. The autocorrelation shows correlations at lags 12 and 24 while the partial autocorrelation shows statistically significant effects at lag 
2, 9-14 and 24. These indicate that the seasonality component remains and the series is not stationary.

```{r}
Box.test(ts1, type="Ljung-Box")
```

The outcome of the Ljung-Box test indicates that we can reject the null hypothesis of series independence. This agrees with our notion that the residuals remain correlated.


Comparative statistics for the time series, fitted model and residuals shows that the model is reasonably close to the time series.

```{r}
fit.df <- data.frame(cbind(ts1, fitted(ts1.fit), ts1.fit$residuals))
class(df)
stargazer(fit.df, type="text",title="Descriptive Statistics", digits=1)
```

The original time series and the ARIMA(1,1,1)(0,1,1)[12] model plot shows that the model captures the behavior of the time series very well and the in-sample fit is good.

```{r}
par(mfrow=c(1,1))
plot.ts(ts1, main="Time Series vs. ARMA(2,2) Model and Residuals",
        ylab="Actual and Estimated Values", xlab="Period",
        col="navy", ylim=c(30,150), xlim=c(0,380), lty=2)
par(new=T)
plot.ts(fitted(ts1.fit), xlab='', ylab='', axes=F, col='blue',
        ylim=c(30,150), xlim=c(0,380), lty=1)
leg.txt <- c("Original Series", "Estimate Series", "Residuals")
legend("topleft",legend=leg.txt,lty=c(1,1,2),
      col=c("navy","blue","green"),
      bty='n', cex=1)
par(new=T)
plot.ts(ts1.fit$residuals, axes=F,xlab='',ylab='',col="green",
        lty=2, pch=1, col.axis='green', ylim=c(-15,50),
        xlim=c(0,380))
axis(side=4,col='green')
mtext("Residuals",side=4,line=2, col='green')
```

We will re-estimate the model by omitting the final 12 time periods and then compare the forecast to the actual values to measure the out-of-sample fit of the model.

```{r}
# fit all but the last 12 periods of the time series
ts1.fit_short <- Arima(ts1[1:360], order=c(2,0,1))
ts1.fit_short
summary(ts1.fit_short)
ts1.fcast_short <- forecast.Arima(ts1.fit_short, h=12)
ts1.fcast_short
summary(ts1.fcast_short$mean)
```

```{r}
par(mfrow=c(1,1))
plot(ts1.fcast_short,
     main='12-Step Ahead Forecast, Original Series and Esitmated Series',
     xlab='', ylab='',
     ylim=c(30,160), xlim=c(0,390), lty=1, col='blue')
par(new=T)
plot.ts(ts1, main="Time Series  ARMA(2,2) Forcast Model and Residuals",
        ylab="Actual and Estimated Values", xlab="Period",
        col="navy", ylim=c(30,160), xlim=c(0,390), lty=2)
leg.txt <- c("Original Series", "Estimate Series", "Residuals")
legend("topleft",legend=leg.txt,lty=c(1,1,2),
      col=c("navy","blue","green"),
      bty='n', cex=1)
par(new=T)
plot.ts(ts1.fit_short$residuals, axes=F,xlab='',ylab='',col="green",
        lty=2, pch=1, col.axis='green', ylim=c(-15,50),
        xlim=c(0,380))
axis(side=4,col='green')
mtext("Residuals",side=4,line=2, col='green')
```

The out-of-series forecast created by estimating a model that omits the final 12 periods of the original time series does not capture the observed values of those final 12 periods. Instead the first half of the final 12 periods are within the statistical boundaries of the forecast but the second half exhibit seasonality effects that are outside the bounds of the forecast.

Forecast Model for 12 time periods. 

It's apparent from the summary statistics and from the forecast graph that the projected forecast for the next 24 periods shows a level trend with a mean value of 146.3. However this forecast doesn't contain the actual observed behavior of the original series; neither the trend nor the seasonality are present.

``` {r}
ts1.fcast <- forecast.Arima(ts1.fit, h=12)
ts1.fcast
summary(ts1.fcast$mean)
```

``` {r}
par(mfrow=c(1,1))
plot(ts1.fcast,
     main='12-Step Ahead Forecast, Original Series and Esitmated Series',
     xlab='Simulated Time Period',
     ylab='Predicted Value and Actual Value',
     ylim=c(30,160), xlim=c(0,390), lty=1, col='blue')
par(new=T)
plot.ts(fitted(ts1.fit), axes=F,
        ylab='', xlab='',
        col="navy", ylim=c(30,160), xlim=c(0,390), lty=2)
legend("topleft",legend=leg.txt,lty=c(1,1,2),col=c("navy","blue","green"),
      bty='n', cex=1)
par(new=T)
plot.ts(ts1.fit$residuals, axes=F,xlab='',ylab='',col="green",
        lty=2, pch=1, col.axis='green', ylim=c(-15,50),xlim=c(0,390) )
axis(side=4,col='green')
mtext("Residuals",side=4,line=2, col='green')
```

