---
title: "W271 - Applied Regression and Time Series Analysis - HW8"
author: "Subhashini R., Lei Yang, Ron Cordell"
date: "April 7, 2016"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

###Build an univariate linear time series model (i.e AR, MA, and ARMA models) using the series in hw08_series.csv.
**Use all the techniques that have been taught so far to build the model, including date examination, data visualization, etc.**
**All the steps to support your final model need to be shown clearly.**
**Show that the assumptions underlying the model are valid.**
**Which model seems most reasonable in terms of satisfying the model’s underling assumption?**
**Evaluate the model performance (both in- and out-of-sample)**
**Pick your “best” models and conduct a 12-step ahead forecast. Discuss your results. Discuss the choice of your metrics to measure “best”.**



```{r}
library(forecast)
# load data
setwd("~/Desktop/W271Data")
x1 <- read.csv('hw08_series.csv', header = T)
```


Let's evaluate some plots
```{r}
par(mfrow=c(2,2))
plot(x1$x, ylab="value", main="hw08_series")
hist(x1$x, col="grey", ylab="value", main="hw08_series")
acf(x1$x, ylab="value", main="hw08_series")
pacf(x1$x, ylab="value", main="hw08_series")
```

according to the plot of ACF and PACF, it is intriguing to build a AR(1) model.

```{r}
ar1.fit <- arima(x1$x, order=c(1,0,0))
  ar1.fit
  summary(ar1.fit)
```

from the coefficient, it's almost a random walk.

Let's check residue for the stationality assumption

```{r}
par(mfrow=c(2,2))
plot(ar1.fit$residuals, ylab="value", main="Residuals")
hist(ar1.fit$residuals, col="grey", ylab="value", main="Residuals")
acf(ar1.fit$residuals, ylab="value", main="Residuals")
pacf(ar1.fit$residuals, ylab="value", main="Residuals")
```

obviously it's not random noise, as the magnitude is increasing in the second half. Although the Box test indicate non-significant results:
```{r}
Box.test(ar1.fit$residuals, type="Ljung-Box")
```

Let's increase the AR order:

```{r}
ar.fit <- ar(x1$x, method="mle") 
  ar.fit
  summary(ar.fit)
```

Let's check residue for the stationality assumption for the new AR model

```{r}
par(mfrow=c(2,2))
plot(ar.fit$resid, ylab="value", main="Residuals")
hist(ar.fit$resid, col="grey", ylab="value", main="Residuals")
acf(ar.fit$resid, ylab="value", main="Residuals", na.action=na.pass)
pacf(ar.fit$resid, ylab="value", main="Residuals", na.action=na.pass)
```

from the plot, it's still not quite random, and the Box test indicates significant results as well. 

Let's check MA model:
```{r}
ma2.fit <- arima(x1$x, order=c(0,0,2))
ma4.fit <- arima(x1$x, order=c(0,0,4))
ma6.fit <- arima(x1$x, order=c(0,0,6))
ma8.fit <- arima(x1$x, order=c(0,0,8))
```

and check the residuals:

```{r}
par(mfrow=c(2,2))
plot(ma2.fit$resid, ylab="value", main="MA(2) Residuals")
plot(ma2.fit$resid, ylab="value", main="MA(4) Residuals")
plot(ma2.fit$resid, ylab="value", main="MA(6) Residuals")
plot(ma2.fit$resid, ylab="value", main="MA(8) Residuals")
```

obviously MA model is not adquate for the series.

Let's turn to ARMA model, and it turns out we are not able to fit ARMA model as the series is not stationary, obviously.

```{r}
arma11 <- arima(x1$x, order=c(1,0,1))
arma21 <- arima(x1$x, order=c(2,0,1))
arma22 <- arima(x1$x, order=c(2,0,2))
#arma23 <- arima(x1$x, order=c(2,0,3))
#arma33 <- arima(x1$x, order=c(3,0,3))
```

Based on the principle of parsimonious, we will use AR(1) as the model, since both the series and the residuals are non-stationary.

```{r}
par(mfrow=c(1,1))
  plot(x1$x, col="red", 
       main="Original vs Estimated Series (AR1)",
       ylab="Values", lty=2)
  lines(fitted(ar1.fit),col="blue")
  leg.txt <- c("Original Series", "Estimated Series")
  legend("topright", legend=leg.txt, lty=c(2,1), 
         col=c("red","blue"), bty='n', cex=1)
```

in-sample prediction fits the original data pretty well.

12-step prediction:
```{r}
ar1.fit.fcast <- forecast.Arima(ar1.fit, 12)
  summary(ar1.fit.fcast)

  plot(ar1.fit.fcast, main="12-Step Ahead Forecast and Original & Estimated Series",
      xlab="Simulated Time Period", ylab="Original, Estimated, and Forecasted Values",
      xlim=c(), lty=2, col="navy")
  lines(fitted(ar1.fit),col="blue")  
```
