---
title: "W271 - Applied Regression and Time Series Analysis - HW7"
author: "Subhashini R., Lei Yang, Ron Cordell"
date: "March 24, 2016"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Question 1:
**1.1 Load hw07_series1.csv**

```{r}
# load data
setwd("~/Desktop/W271Data")
x1 <- read.csv('hw07_series1.csv', header = T)
```


**1.2 Describe the basic structure of the data and provide summary statistics of the series**

```{r}
str(x1)
head(x1)
summary(x1$X10.01)
```


**1.3 Plot histogram and time-series plot of the series. Describe the patterns exhibited in histograrm and time-series plot. For time series analysis, is it sufficient to use only histogram to describe a series?**

```{r, warning=FALSE}
par(mfrow=c(2,2))
    hist(x1$X10.01)
    plot.ts(x1, type='l',
         main="A time series with 74 observations",
         ylab="Values", xlab="Time Period")
    acf(x1, main="Autocorrelation Diagram")
    pacf(x1, main="Partial Autocorrelation Diagram")
```

the histogram shows some negative skew. The time-series plot is increasing at the beginning, then drops a bit before another cycle of increase and drop. With limited sample size, it's hard to conclude if there is any trend or seasonal pattern in the data. Histogram alone is **not** enough, as it doesn't show the temporal information in the data, thus we can tell the dynamics of the series.

**1.4 Plot the ACF and PACF of the series. Describe the patterns exhibited in the ACF and PACF. **

The ACF has shown a gradual decay with positive value, and the PACF drops significantly after 2 lags. Based on these observation, it is like the series is an $AR(1)$ model.

**1.5 Estimate the series using the ar() function.**

```{r}
ar.1 <- ar(x1$X10.01, method="mle") 
ar.1
```

**1.6 Report the estimated AR parameters, the order of the model, and standard errors.**

The order the model is **``r sprintf("%d", ar.1$order)``**, and standard errors is **``r sprintf("%.4f", sqrt(ar.1$var.pred))``**.


# Question 2:

**2.1 Simulate a time series of lenght 100 for the following model. Name the series $x$.** 

$$x_t = \frac{5}{6} x_{t-1} - \frac{1}{6} x_{t-2} + \omega_t$$

```{r}
set.seed(898)
    n <- 100
    x <- w <- rnorm(n)
    for (t in 3:n) x[t] <- 5*x[t-1]/6 - x[t-2]/6 + w[t] # a zero-mean AR(2) process
    summary(x)    
```

**2.2 Plot the correlogram and partial correlogram for the simulated series. Comments on the plots.**

```{r}
par(mfrow=c(2,2))
    hist(x)
    plot.ts(x, type='l',
         main="A Simulated AR(2) Model",
         ylab="Values", xlab="Time Period")
    acf(x, main="Autocorrelation Diagram")
    pacf(x, main="Partial Autocorrelation Diagram")
```

There isn't any trend or seasonal pattern from the series, both ACF and PACF oscillate. However, the PACF drops significantly just after 1st leg, which indicates an AR order of $1$.

**2.3 Estimate an AR model for this simulated series. Report the estimated AR parameters, standard erros, and the order of the AR model.**

```{r}
ar.2 <- ar(x, method="mle") 
ar.2
```

The order the model is **``r sprintf("%d", ar.2$order)``**, and standard errors is **``r sprintf("%.4f", sqrt(ar.2$var.pred))``**.

**2.4 Construct a 95% confidence intervals for the parameter estimates of the estimated model. Do the “true” model parameters fall within the confidence intervals? Explain the 95% confidence intervals in this context.**

The 95% confidence intervals for the estimated paramter is:

```{r}
ar.2$ar + c(-2,2)*sqrt(ar.2$asy.var)
```

the "true" model does not have the same order with estimated model.


**2.5 Is the estimated model stationary or non-stationary?**

For the estimated model:

```{r}
polyroot(c(1,-ar.2$ar))
```

the single root is bigger than zero, thus the model is stationary.

**2.6 Plot the correlogram of the residuals of the estimated model. Comment on the plot.**

Check the residual:

```{r}
par(mfrow=c(2,2))
    hist(ar.2$resid)
    plot.ts(ar.2$resid, type='l',
         main="Residual of the estimated Model",
         ylab="Values", xlab="Time Period")
    acf(ar.2$resid, main="Residual ACF", na.action = na.omit)
    pacf(ar.2$resid, main="Residual PACF", na.action = na.omit)
```

From both ACF and PACF the residuals don't have any correlation and demonstrate random pattern.